---
title: "Relationship to base and plyr functions"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(tidyverse)
library(gapminder)
```


  * List relevant base and plyr functions. Create a base <--> purrr, base <--> plyr map.
  * Point out general reasons to consider purrr.
  * Point out specific gotchas or FAQs that purrr avoids or answers.

### Base functions

`lapply()` <--> `purrr::map()`

`sapply()`, fine for interactive use, but unpredictability hard to program around

`vapply()`, some similarities with `map_chr()`, `map_lgl()`, etc. Touted as the base option for programming, but still unpredictable.

Example gotcha from real life. Using `vapply()` inside a function. `n` can vary depending on input. But `vapply()`'s simplifying behavior means it will return 2D or 1D object, depending on `n`. It has no way to say `drop = FALSE`.

<div class = "row">
<div class = "col-md-6">
```{r}
f <- function(x, n) rep(x, n)
n <- 3
vapply(c("a", "b"), f, character(n), n = n)
n <- 1
vapply(c("a", "b"), f, character(n), n = n)
```
</div>
<div class = "col-md-6">
```{r}
f <- function(x, n) rep(x, n)
map(c("a", "b"), f, n = 3) %>% lift_dl(cbind)(.)
map(c("a", "b"), f, n = 1) %>% lift_dl(cbind)(.)
```
</div>
</div>

`mapply()`, similar to `map2()`, `pmap()`

`tapply()`, `aggregate()`

Possible format for some comparisons

<div class = "row">
<div class = "col-md-6">
```{r}
aggregate(breaks ~ wool + tension,
          data = warpbreaks, mean)
```
</div>
<div class = "col-md-6">
```{r}
warpbreaks %>% 
  group_by(wool, tension) %>% 
  summarise(mean = mean(breaks), var = var(breaks))
```
</div>
</div>




