---
title: "Sample from groups, n varies by group"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A challenge [tweeted by Hilary
Parker](https://twitter.com/hspter/status/739886244692295680), paraphrased:

> How do you sample from groups, with a different sample size for each group?

Illustrated with the iris data.

  * `Species` = groups.
  * Sample from the 3 `Species` with 3 different sample sizes.
  
How fits the template:

DRAW A SAMPLE for each PAIR OF (SPECIES DATA, SPECIES SAMPLE SIZE)

How to prepare the data? I need a data frame with

  * One row per `Species`
  * A variable of `Species`-specific sample sizes
  * A variable of "Species data", whatever that means.
    - Actually we know what that is: a variable of `Species`-specific data frames. A list-column!

We need a *nested data frame*.

```{r}
suppressMessages(library(dplyr))
suppressMessages(library(purrr))
library(tidyr)
set.seed(4561)

(nested_iris <- iris %>%
    group_by(Species) %>%   # prep for work by Species
    nest() %>%              # --> one row per Species
    mutate(n = c(2, 5, 3))) # add sample sizes
```

Draw the samples.

  * `purrr::map2()` is good since we want to operate on 2 things `(data = DATA FOR ONE SPECIES, n = SAMPLE SIZE)`.
  * We've already got `data = DATA FOR ONE SPECIES` and `n = SAMPLE SIZE` as variables in our data frame.
  * Drop them in as inputs 1 and 2 to `dplyr::sample_n(tbl, size)`.
  * Accept whatever comes back as a new list-column in the data frame, i.e. use `dplyr::mutate()`. Be brave and deal with it.

```{r}
(sampled_iris <- nested_iris %>%
  mutate(samp = map2(data, n, sample_n)))
```

What came back? More `Species`-specific data frames.

We are in that uncomfortable intermediate state, with two list-columns: the original `data` and the sampled data, `samp`. Let's get back to a normal data frame!

  * Keep only `Species` and `samp` variables.
  * Unnest, which essentially rowbinds the data frames in `samp` and replicates `Species` as necessary.

```{r}
sampled_iris %>% 
  select(Species, samp) %>%
  unnest()
```

Again, from the top, with no exposition:

```{r}
iris %>%
  group_by(Species) %>% 
  nest() %>%            
  mutate(n = c(2, 5, 3)) %>% 
  mutate(samp = map2(data, n, sample_n)) %>% 
  select(Species, samp) %>%
  unnest()
```

Base R solution, with marginal comments about what bugs me, especially as someone who teaches R:

```{r}
split_iris <- split(iris, iris$Species) # why can't Species be found in iris?
                                        # where else would it be found?
str(split_iris)                         # split_iris ~= nested_iris[["data"]]
(n <- c(2, 5, 3))                       # Species data and n are only 'in sync'
                                        # due to my discipline / care
                                        # not locked safely into a data frame
(group_sizes <- vapply(split_iris, nrow, integer(1))) # also floating free
(sampled_obs <- mapply(sample, group_sizes, n)) # I'm floating free too!
get_rows <- function(df, rows) df[rows, , drop = FALSE] # custom function
                                        # drop = FALSE required to avoid
                                        # nasty surprise in case of n = 1
(sampled_iris <-                        # god help you if forget SIMPLIFY = FALSE
    mapply(get_rows, split_iris, sampled_obs, SIMPLIFY = FALSE))
do.call(rbind, sampled_iris)            # :( do.call()
```

IMO the base R solution requires much greater facility with R programming and data structures to get it right. It feels more like programming than data analysis.

