---
title: "Making a data frame from a list from GitHub API"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

One of the best uses for `purrr`'s `map()` functions is for digging information out of a nested list and creating a neat data frame. Where do these lists come from? Often as JSON or XML from an API.

Here we'll explore some lists that come from calling the [GitHub API](https://developer.github.com/v3/).

### The `gh` package and getting a GitHub user

```{r}
library(purrr)
## devtools::install_github("gaborcsardi/gh")
library(gh)
library(listviewer)
suppressMessages(library(dplyr))
```

Use `gh::gh()` (not explained here) to get info on a specific GitHub user.

*come back with code to grab this rds from github*

```{r include = FALSE, eval = FALSE}
gabor <- gh("/users/:username", username = "gaborcsardi")
saveRDS(gabor, "gabor.rds")

unames <-
  c("gaborcsardi", "jennybc", "masalmon", "juliasilge", "leeper", "jtleek")
users <- map(unames, ~ gh("/users/:username", username = .x))

## why is there more info for me than anyone else?
nms <- users %>%
  map(names)
setdiff(nms[[2]], nms[-2] %>% reduce(union))
# [1] "private_gists"       "total_private_repos" "owned_private_repos"
# [4] "disk_usage"          "collaborators"       "plan"               
# I guess this doesn't matter
saveRDS(users, "github_users.rds")
```

```{r}
# gabor <- gh("/users/:username", username = "gaborcsardi")
gabor <- readRDS("gabor.rds")
str(gabor, give.attr = FALSE)
```

### Getting several GitHub users

What if we want information on several GitHub users? The `gh::gh()` function is not vectorized but we can use `map()` to iteratively make `gh::gh()` calls. This will yield a list with info just like that above, but for several users.

*come back with code to grab this rds from github*

```{r}
unames <-
  c("gaborcsardi", "jennybc", "masalmon", "juliasilge", "leeper", "jtleek")
# users <- map(unames, ~ gh("/users/:username", username = .x))
users <- readRDS("github_users.rds")
jsonedit(setNames(users, unames))
```

<!--
To have the interactive list viewing experience, visit <http://rpubs.com/jennybc/github-users>.
-->

### Character and integer vector shortcut

Now we want to reach into each user's list and starting pulling info out to make variables in a data frame. This showcases a specific shortcut for making the `.f` function that `map()` will apply.

Provide character or integer input as the `.f` function and it's as if you provided a function like `function(x) x[["y"]]` or `function(i) x[[i]]`. You can index at deeper levels with a character or integer vector, but our current example doesn't support this.

```{r}
users %>% 
  map_chr("name")

## the first element of each user's list is their username
users %>% 
  map_chr(1)
```

### Multiple values at same level of hierarchy

A trick for retrieving multiple values at the same level of the hierarchy is to map the function `[` and provide a character or integer vector as an additional argument.

```{r}
x <- users %>%
  map(`[`, c("login", "name", "id", "location"))
str(x[[1]])

y <- users %>% 
  map(`[`, 1:2)
str(y[[2]])
```

What's really awesome is to stack up these per-users lists row-wise and make a data frame. We use the special mapping function `map_df()` to signal that we want a data frame back.

```{r}
users %>%
  map_df(`[`, c("login", "name", "id", "location"))
```

Another, more cumbersome but safer way to get this same data frame is to form variables explicitly like so:

```{r}
users %>% {
  data_frame(
    login = map_chr(., "login"),
    name = map_chr(., "name"),
    id = map_int(., "id"),
    location = map_chr(., "location")
  )
}
```

### Get repos by user

*under development*

```{r}
library(tibble)
# repos <- unames %>% 
#   map(~ gh("/users/:user/repos", user = .x, .limit = Inf))
# saveRDS(repos, "github_repos.rds")
repos <- readRDS("github_repos.rds")
(rdf <- repos %>%
    set_names(unames) %>% 
    enframe(value = "gh_repos"))
jsonedit(set_names(repos, unames))
```

Make user-specific data frames, one row per repo. Then `unnest()` the enclosing data frame to get one row per repo.

```{r}
rdf %>% 
  mutate(
    #n_repos = map_int(gh_repos, length),
    repo_info = gh_repos %>%
      map(. %>% map_df(`[`, c("name", "fork", "open_issues")))
  ) %>% 
  select(-gh_repos) %>% 
  tidyr::unnest()
```

