---
title: "Explore the example lists: Wes Anderson, Game of Thrones, GitHub"
comment: "*how to get to know a list*"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = TRUE)
```

## Load packages

Load purrr and repurrrsive, which contains recursive list examples.

```{r}
library(purrr)
library(repurrrsive)
```

## Inspect and explore

List inspection is very important and also fairly miserable. Before you can apply a function to every element of a list, you'd better understand the list!

You need to develop a toolkit for list inspection. Be on the look out for:

  * What is the length of the list?
  * Are the components homogeneous, i.e. do they have the same overall structure, albeit containing different data?
  * Note the length, names, and types of the constituent objects.
  
> I have no idea what's in this list or what its structure is! Please send help.

Understand this is **situation normal**, especially when your list comes from querying a poorly documented API. This is often true even when your list has been created completely within R. How many of us perfectly understand the structure of a fitted linear model object? You just have to embark on a voyage of discovery and figure out what's in there. Happy trails.

### Indexing, review

Rembmer, there are 3 ways to pull elements out of a list:

  * The `$` operator. Extracts a single element by name. Name can be unquoted, if syntactic.
    ```{r}
    x <- list(a = "a", b = 2)
    x$a
    x$b
    ```
  * `[[` a.k.a. double square bracket. Extracts a single element by name or position. Name must be quoted, if provided directly. Name or position can also be stored in a variable.
    ```{r}
    x <- list(a = "a", b = 2)
    x[["a"]]
    x[[2]]
    nm <- "a"
    x[[nm]]
    i <- 2
    x[[i]]
    ```
  * `[` a.k.a. single square bracket. Regular vector indexing. For a list input, this always returns a list!
    ```{r}
    x <- list(a = "a", b = 2)
    x["a"]
    x[c("a", "b")]
    x[c(FALSE, TRUE)]
    ```

### `str()`

`str()` can help with basic list inspection, although it's still rather frustrating. Learn to love the `max.level` and `list.len` arguments. You can use them to keep the output of `str()` down to a manageable volume.

Once you begin to suspect or trust that your list is homogeneous, i.e. consists of sub-lists with similar structure, it's often a good idea to do an in-depth study of a single element. In general, remember you can combine list inspection via `str(..., list.len = x, max.level = y)` with single `[` and double `[[` square bracket indexing.

The repurrrsive package provides examples of lists. We explore them below, to lay the groundwork for other lessons, and to demonstrate list inspection strategies.

### listviewer and RStudio's Object Explorer

 The RStudio IDE (v1.1 and higher) offers an [Object Explorer](https://blog.rstudio.com/2017/08/22/rstudio-v1-1-preview-object-explorer/) that provides interactive inspection and code generation tools for hierarchical objects, such as lists. You can invoke it via the GUI or in code as `View(YOUR_UGLY_LIST)`.
 
However, that won't help you expose list exploration in something like this website. I am using the [listviewer](https://CRAN.R-project.org/package=listviewer) package to do this below. It allows you to expose list exploration in a rendered `.Rmd` document.
To replicate this experience locally, call, e.g., `listviewer::jsonedit(got_chars, mode = "view")`.


```{r}
library(listviewer)
```

## Wes Anderson color palettes

`wesanderson` is a simple list containing color palettes from the [wesanderson package](https://cran.r-project.org/package=wesanderson). Each component is a palette, named after a movie, and contains a character vector of colors as hexadecimal triplets.

```{r}
str(wesanderson)
```

### Explore `wesanderson`

```{r echo = FALSE}
jsonedit(wesanderson, mode = "view", elementId = "wesanderson")
```

You can get a similar experience in RStudio via `View(wesanderson)`.

## Game of Thrones POV characters

`got_chars` is a list with information on the `r length(got_chars)` point-of-view characters from the first five books in the Song of Ice and Fire series by George R. R. Martin. Retrieved from [An API Of Ice And Fire](https://anapioficeandfire.com). Each component corresponds to one character and contains `r length(got_chars[[1]])` components which are named atomic vectors of various lengths and types.

```{r}
str(got_chars, list.len = 3)
str(got_chars[[1]], list.len = 8)
```

### Explore `got_chars`

```{r echo = FALSE}
jsonedit(number_unnamed(got_chars), mode = "view", elementId = "got_chars")
```

You can get a similar experience in RStudio via `View(got_chars)`.

## GitHub users and repositories

`gh_users` is a list with information on 6 GitHub users. `gh_repos` is a nested list, also of length 6, where each component is another list with information on up to 30 of that user's repositories. Retrieved from the [GitHub API](https://developer.github.com/v3/).

```{r}
str(gh_users, max.level = 1)
```

### Explore `gh_users`

```{r echo = FALSE}
jsonedit(number_unnamed(gh_users), mode = "view", elementId = "gh_users")
```

You can get a similar experience in RStudio via `View(gh_users)`.

### Explore `gh_repos`

```{r echo = FALSE}
jsonedit(number_unnamed(gh_repos), mode = "view", elementId = "gh_repos")
```

You can get a similar experience in RStudio via `View(gh_repos)`.

#### Exercises

1. Read the documentation on `str()`. What does `max.level` control? Apply `str()` to `wesanderson` and/or `got_chars` and experiment with `max.level = 0`, `max.level = 1`, and `max.level = 2`. Which will you use in practice with deeply nested lists?
1. What does the `list.len` argument of `str()` control? What is its default value? Call `str()` on `got_chars` and then on a single component of `got_chars` with `list.len` set to a value much smaller than the default. What range of values do you think you'll use in real life?
1. Call `str()` on `got_chars`, specifying both `max.level` and `list.len`.
1. Call `str()` on the first element of `got_chars`, i.e. the first Game of Thrones character. Use what you've learned to pick an appropriate combination of `max.level` and `list.len`.
