---
title: "Explore the example lists: Wes Anderson, Game of Thrones, GitHub"
comment: "*how to get to know a list*"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = TRUE)
```

## Load packages

Load purrr and repurrrsive, which contains recursive list examples.

```{r}
library(purrr)
library(repurrrsive)
```

List inspection is very important and also fairly miserable. Below I offer interactive widgets for our example lists. If you wish to do the same locally, install the [listviewer](https://CRAN.R-project.org/package=listviewer) package.

```{r}
library(listviewer)
```

## Inspect and explore

Before you can apply a function to every element of a list, you'd better understand the list!

You need to develop a toolkit for list inspection. Be on the look out for:

  * What is the length of the list?
  * Are the components homogeneous, i.e. do they have the same overall structure, albeit containing different data?
  * Note the length, names, and types of the constituent objects.
  
> I have no idea what's in this list or what its structure is! Please send help.

Understand this is **situation normal**, especially when your list comes from querying a poorly documented API. This is often true even when your list has been created completely within R. How many of us perfectly understand the structure of a fitted linear model object? You just have to embark on a voyage of discovery and figure out what's in there. Happy trails.

### `str()`

`str()` can help with basic list inspection, although it's still rather frustrating. Learn to love the `max.level` and `list.len` arguments. You can use them to keep the output of `str()` down to a manageable volume.

Once you begin to suspect or trust that your list is homogeneous, i.e. consists of sub-lists with similar structure, it's often a good idea to do an in-depth study of a single element. In general, remember you can combine list inspection via `str(..., list.len = x, max.level = y)` with single `[` and double `[[` square bracket indexing.

The repurrrsive package provides examples of lists. We explore them below, to lay the groundwork for other lessons, and to demonstrate list inspection strategies.

### listviewer

It's often more helpful to explore such lists with an interactive widget, which we include below for all examples. If you installed the listviewer package, replicate this experience locally via `listviewer::jsonedit(got_chars, mode = "view")`.

## Wes Anderson color palettes

`wesanderson` is a simple list containing color palettes from the [wesanderson package](https://cran.r-project.org/package=wesanderson). Each component is a palette, named after a movie, and contains a character vector of colors as hexadecimal triplets.

```{r}
str(wesanderson)
```

### listview `wesanderson`

```{r echo = FALSE}
jsonedit(wesanderson, mode = "view", elementId = "wesanderson")
```

## Game of Thrones POV characters

`got_chars` is a list with information on the `r length(got_chars)` point-of-view characters from the first five books in the Song of Ice and Fire series by George R. R. Martin. Retrieved from [An API Of Ice And Fire](https://anapioficeandfire.com). Each component corresponds to one character and contains `r length(got_chars[[1]])` components which are named atomic vectors of various lengths and types.

```{r}
str(got_chars, list.len = 3)
str(got_chars[[1]], list.len = 8)
```

### listview `got_chars`

```{r echo = FALSE}
jsonedit(number_unnamed(got_chars), mode = "view", elementId = "got_chars")
```

## GitHub users and repositories

`gh_users` is a list with information on 6 GitHub users. `gh_repos` is a nested list, also of length 6, where each component is another list with information on up to 30 of that user's repositories. Retrieved from the [GitHub API](https://developer.github.com/v3/).

```{r}
str(gh_users, max.level = 1)
```

### listview `gh_users`

```{r echo = FALSE}
jsonedit(number_unnamed(gh_users), mode = "view", elementId = "gh_users")
```

### listview `gh_repos`

```{r echo = FALSE}
jsonedit(number_unnamed(gh_repos), mode = "view", elementId = "gh_repos")
```

#### Exercises

1. Read the documentation on `str()`. What does `max.level` control? Apply `str()` to `wesanderson` and/or `got_chars` and experiment with `max.level = 0`, `max.level = 1`, and `max.level = 2`. Which will you use in practice with deeply nested lists?
1. What does the `list.len` argument of `str()` control? What is its default value? Call `str()` on `got_chars` and then on a single component of `got_chars` with `list.len` set to a value much smaller than the default. What range of values do you think you'll use in real life?
1. Call `str()` on `got_chars`, specifying both `max.level` and `list.len`.
1. Call `str()` on the first element of `got_chars`, i.e. the first Game of Thrones character. Use what you've learned to pick an appropriate combination of `max.level` and `list.len`.
