---
title: "Simplifying data from a list from GitHub API"
comment: "*suitable for a hands-on activity*"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

One of the best uses for `purrr`'s `map()` functions is for digging information out of a non-rectangular data structures and creating a neat data frame. Where do these awkward objects come from? Often as JSON or XML from an API, which has frequently been pre-converted to an R list.

Here we explore some lists that come from calling the [GitHub API](https://developer.github.com/v3/). Interactive exploration of these lists is made possible by the [`listviewer` package](https://github.com/timelyportfolio/listviewer).

This example also shows how character and integer vectors can be provided as the function `.f` for the `map()` family functions. An extractor function is created behind the scenes, providing a concise way to extract by name or position.

```{r}
library(purrr)
suppressMessages(library(jsonlite))
library(listviewer)
library(tibble)
suppressMessages(library(dplyr))
```

### Get several GitHub users

```{r include = FALSE, eval = FALSE}
unames <-
  c("gaborcsardi", "jennybc", "masalmon", "juliasilge", "leeper", "jtleek")
urls <- paste0("https://api.github.com/users/", unames)
users <- urls %>%
  map(fromJSON)
users %>% 
  toJSON() %>% 
  prettify() %>%
  writeLines("github_users.json")
```

Here is information on six GitHub users, pre-stored as JSON:  
<https://github.com/jennybc/purrr-tutorial/raw/gh-pages/github_users.json>. Use the code below

```{r}
users <- fromJSON("github_users.json", simplifyDataFrame = FALSE)
str(users, max.level = 1)
str(users[[1]], list.len = 6)
```

`users` is *a nested list*:

  * one element per GitHub user
  * each element is, in turn, a list with info on the user

Explore it interactively:

```{r echo = FALSE}
jsonedit(users)
```

<!--
To have the interactive list viewing experience, visit <http://rpubs.com/jennybc/github-users>.
-->

### Shortcuts to extract elements by name or position

Who are these people?

Reach into each user's list and pull elements out, such as "name".

Shortcut for making the `.f` function that `map()` will apply:

  * provide "TEXT" to extract the element named "TEXT"
  * equivalent to `function(x) x[["TEXT"]]`

```{r}
users %>% 
  map_chr("name")

## equivalent to the more verbose
# users %>%
#   map_chr(function(x) x[["name"]])

(unames <- users %>%
    map_chr("login"))
```

Or extract by position:

  * provide `i` to extract the `i`-th element
  * equivalent to `function(x) x[[i]]`

```{r}
users %>% 
  map_chr(1)

## equivalent to the more verbose
# users %>%
#   map_chr(function(x) x[[1]])
```

### Multiple values at same level of hierarchy

A trick for retrieving multiple values at the same level of the hierarchy is to map the function `[` and provide a character or integer vector as an additional argument.

```{r}
x <- users %>%
  map(`[`, c("login", "name", "id", "location"))
str(x[[1]])

## if you find this ugly, try magrittr::extract()
x <- users %>%
  map(magrittr::extract, c("login", "name", "id", "location"))

y <- users %>% 
  map(`[`, 1:2)
str(y[[2]])
```

What's really awesome is to stack up these per-user lists row-wise and make a data frame. We use the special mapping function `map_df()` to signal that we want a data frame back.

```{r}
users %>%
  map_df(`[`, c("login", "name", "id", "location"))
```

Another, more cumbersome but safer way to get this same data frame is to form variables explicitly like so:

```{r}
users %>% {
  tibble(
    login = map_chr(., "login"),
    name = map_chr(., "name"),
    id = map_int(., "id"),
    location = map_chr(., "location")
  )
}
```

### Repositiories for each user

```{r include = FALSE, eval = FALSE}
urls <- paste0("https://api.github.com/users/", unames, "/repos")
repos <- urls %>% map(fromJSON)
writeLines(toJSON(repos), "github_repos.json")
```

The list above has one level of nesting, but it's common to have even more.

`repos` is a list with:

  * one element per GitHub user
  * each element is another list of that user's repositories (actually, just the first 30)
  * several of those list elements are, again, a list

Download this:  
<https://github.com/jennybc/purrr-tutorial/raw/gh-pages/github_repos.json>

```{r}
repos <- fromJSON("github_repos.json", simplifyDataFrame = FALSE)
str(repos, max.level = 1)
```

Explore it interactively:

```{r echo = FALSE}
jsonedit(repos)
```

### Revisit shortcuts to extract elements by name or position

Now we can use the indexing shortcuts in a more complicated setting. Instead of providing a single name or position, we use a vector:

  * the `j`-th element addresses the `j`-th level of the hierarchy
  
It's easiest to see in a concrete example. We get the full name (element 3) of the first repository listed for each user.

```{r}
repos %>%
  map_chr(c(1, 3))
## TO DO: I would really prefer a character example :(
```

### Organizing an analysis in a data frame

*under development*

Mission: get a data frame with one row per repository, with varibles identifying which GitHub user owns it and the repository name, etc.

Step 1: Put the repository list into a data frame, along with identifying GitHub usernames.

```{r}
(udf <- repos %>%
    set_names(unames) %>% 
    enframe("username", "gh_repos"))
```

*go out in blaze of glory ... or rather: develop this in steps*

```{r}
rdf <- udf %>% 
  mutate(
    #n_repos = map_int(gh_repos, length),
    repo_info = gh_repos %>%
      map(. %>% map_df(`[`, c("name", "fork", "open_issues")))
  ) %>% 
  select(-gh_repos) %>% 
  tidyr::unnest()
rdf %>% 
  filter(!fork) %>% 
  select(-fork) %>% 
  group_by(username) %>%
  arrange(username, desc(open_issues)) %>%
  top_n(3)
```

### Appendix

If you just wanted to solve this problem, you could let [`jsonlite`](https://github.com/jeroenooms/jsonlite) simplify the JSON for you. Other packages for list handling include [`listless`](https://bitbucket.org/graumannlabtools/listless), [`rlist`](https://renkun.me/rlist/).
