---
title: "List columns"
comment: "*creating, managing, and eliminating list-columns*"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Data frames are a fantastic data structure for data analysis. We usually think of them as a data receptacle for several atomic vectors with a common length and with a notion of "observation", i.e. the i-th value of each atomic vector is related to all the other i-th values.

But data frame are not limited to atomic vectors. They can host general vectors, i.e. *lists* as well. This is what I call a **list-column**.

List-columns and the data frame that hosts them require some special handling. In particular, it is highly advantageous if the data frame is a [tibble](https://github.com/tidyverse/tibble#readme), which anticipates list-columns. To work comfortably with list-columns, you need to develop techniques to:

  * **Inspect**. What have I created?
  * **Index**. How do I pull out specific bits by name or position?
  * **Compute**. How do I operate on my list-column to make another vector or list-column?
  * **Simplify**. How do I get rid of this list-column and back to a normal data frame?
  
The purrr package and all the techniques depicted in the other lessons come into heavy play here. This is a collection of worked examples that show these techniques applied specifically to list-columns.

## Regex and Trump tweets

### Load packages

```{r message = FALSE}
library(tidyverse)
library(stringr)
library(lubridate)
library(here) ## install.packages("krlmlr/here")
```

### Bring tweets in

Working with the same 7 tweets as [Trump Android words](ls08_trump-tweets.html) lesson. Go there for the rationale for choosing these 7 tweets.

```{r}
tb_raw <- read_csv(here("talks", "trump-tweets.csv"))
```

### Create a list-column of Trump Android words

Clean a variable and create a list-column:

  * `source` comes in an unfriendly form. Simplify to convey if tweet came from Android or iPhone.
  * `twords` are what we'll call the "Trump Android words". See [Trump Android words](ls08_trump-tweets.html) lesson for backstory. **This is a list-column!**

```{r}
source_regex <- "android|iphone"
tword_regex <- "badly|crazy|weak|spent|strong|dumb|joke|guns|funny|dead"

tb <- tb_raw %>%
  mutate(source = str_extract(source, source_regex),
         twords = str_extract_all(tweet, tword_regex))
```

### Derive new variables

Add variables, two of which are based on the `twords` list-column.

 * `n`: How many twords are in the tweet?
 * `hour`: At which hour of the day was the tweet?
 * `start`: Start character of each tword.
 
```{r}
tb <- tb %>%
  mutate(n = lengths(twords),
         hour = hour(created),
         start = gregexpr(tword_regex, tweet))
```

```{r include = FALSE}
# another possibilty that would require more processing
# so less useful for a talk example
# but more useful IRL:
# str_locate_all(tweet, tword_regex))
```

### Use regular data manipulation toolkit

Let's isolate tweets created before 2pm, containing 1 or 2 twords, in which there's an tword that starts within the first 30 characters.

```{r}
tb %>%
  filter(hour < 14,
         between(n, 1, 2),
         between(map_int(start, min), 0, 30))
```

Let's isolate tweets that contain both the twords "strong" and "weak".

```{r}
tb %>%
  filter(map_lgl(twords, ~ all(c("strong", "weak") %in% .x)))
```

## JSON from an API and Game of Thrones

### Load packages

```{r}
library(repurrrsive)
library(tidyverse)
library(httr)
library(stringr)
library(here)
```

### Call the API of Ice and Fire

Here's a simplified version of how we obtained the data on the Game of Thrones POV characters. This data appears as a more processed list in the [repurrrsive](https://cran.r-project.org/package=repurrrsive) package.

  * Get character IDs from repurrrsive. *cheating a little, humor me*
  * Put IDs and character names in a tibble.

```{r}
pov <- set_names(map_int(got_chars, "id"),
                 map_chr(got_chars, "name"))
tail(pov, 5)
ice <- pov %>%
  enframe(value = "id")
ice
```

Request info for each character and store what comes back -- whatever that may be -- in the list-column `stuff`.

```{r}
ice_and_fire_url <- "https://anapioficeandfire.com/"
if (file.exists(here("talks", "ice.rds"))) {
  ice <- readRDS(here("talks", "ice.rds"))
} else {
  ice <- ice %>%
    mutate(
      response = map(id,
                     ~ GET(ice_and_fire_url,
                           path = c("api", "characters", .x))),
      stuff = map(response, ~ content(.x, as = "parsed",
                                      simplifyVector = TRUE))
    ) %>%
    select(-id, -response)
  saveRDS(ice, here("talks", "ice.rds"))
}
ice
```

Let's switch to a nicer version of `ice`, based on the list in repurrrsive, because it already has books and houses replaced with names instead of URLs.

```{r}
ice2 <- tibble(
  name = map_chr(got_chars, "name"),
  stuff = got_chars
)
ice2
```

Inspect the list-column.
```{r}
str(ice2$stuff[[9]], max.level = 1)
# if (interactive()) {
#   listviewer::jsonedit(ice2$stuff[[2]], mode = "view", width = 500, height = 530)
# }
```

### Use regular data manipulation toolkit

Form a sentence of the form "NAME was born AT THIS TIME, IN THIS PLACE" by digging info out of the `stuff` list-column and placing into a string template. No list-columns left!

```{r}
template <- "${name} was born ${born}."
birth_announcements <- ice2 %>%
  mutate(birth = map_chr(stuff, str_interp, string = template)) %>%
  select(-stuff)
birth_announcements
```

Extract each character's house allegiances. Keep only those with more than one allegiance. Then unnest to explode the `houses` list-column and get a tibble with one row per character * house combination. No list-columns left!

```{r}
allegiances <- ice2 %>%
  transmute(name,
            houses = map(stuff, "allegiances")) %>%
  filter(lengths(houses) > 1) %>%
  unnest(houses)
allegiances
```

## Aliases and allegiances of Game of Thrones characters

### Load packages

```{r}
library(tidyverse)
library(repurrrsive)
library(stringr)
```

## Lists as variables in a data frame

One row per GoT character. List columns for aliases and allegiances.

```{r}
x <- tibble(
  name = got_chars %>% map_chr("name"),
  aliases = got_chars %>% map("aliases"),
  allegiances = got_chars %>% map("allegiances")
)
x
#View(x)
```

What if we only care about characters with a "Lannister" alliance? Practice operating on a list-column.

```{r}
x %>%
  mutate(lannister = map(allegiances, str_detect, pattern = "Lannister"),
         lannister = map_lgl(lannister, any))
```

Keep only the Lannisters and Starks allegiances. You can use `filter()` with list-columns, but you will need to `map()` to list-ize your operation. Once I've got the characters I want, I drop `allegiances` and use `unnest()` to get back to a simple data frame with no list columns.

```{r}
x %>%
  filter(allegiances %>%
           map(str_detect, "Lannister|Stark") %>%
           map_lgl(any)) %>%
  select(-allegiances) %>%
  filter(lengths(aliases) > 0) %>%
  unnest(aliases) %>% 
  print(n = Inf)
```

```{r eval = FALSE, include = FALSE}
x_base <- data.frame(
  name = vapply(got_chars, `[[`, character(1), "name"),
  aliases = I(lapply(got_chars, `[[`, "aliases")),
  allegiances = I(lapply(got_chars, `[[`, "allegiances"))
)
keep1 <- vapply(x_base$allegiances, function(y) any(grepl("Lannister|Stark", y)), logical(1))
x_base <- x_base[keep1, ]
x_base$allegiances <- NULL
x_base
data.frame(
  name = rep(x_base$name, lengths(x_base$aliases)),
  aliases = unlist(x_base$aliases)
)
```

## Nested data frame, modelling, and Gapminder

Another version of this same example is here:

<http://r4ds.had.co.nz/many-models.html>

*mostly code at this point, more words needed*

### Load packages

```{r}
library(tidyverse)
library(gapminder)
library(broom)
```

### Hello, again, Gapminder

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp, group = country)) +
    geom_line(alpha = 1/3)
```

What if we fit a line to each country?

```{r}
gapminder %>%
  ggplot(aes(year, lifeExp, group = country)) +
  geom_line(stat = "smooth", method = "lm",
            alpha = 1/3, se = FALSE, colour = "black")
```

What if you actually want those fits? To access estimates, p-values, etc. In that case, you need to fit them yourself. How to do that?

  * Put the variables needed for country-specific models into nested dataframe. In a **list-column**!
  * Use the usual "map inside mutate", possibly with the broom package, to pull interesting information out of the 142 fitted linear models.
  
### Nested data frame
  
Nest the data frames, i.e. get one meta-row per country:

```{r}
gap_nested <- gapminder %>%
  group_by(country) %>%
  nest()
gap_nested
gap_nested$data[[1]]
```

*Compare/contrast to a data frame grouped by country (dplyr-style) or split on country (base)*.

### Fit models, extract results

Fit a model for each country.

```{r}
gap_fits <- gap_nested %>%
  mutate(fit = map(data, ~ lm(lifeExp ~ year, data = .x)))
```

Look at one fitted model, for concreteness.

```{r}
gap_fits %>% tail(3)
canada <- which(gap_fits$country == "Canada")
summary(gap_fits$fit[[canada]])
```

Let's get all the r-squared values!

```{r}
gap_fits %>%
  mutate(rsq = map_dbl(fit, ~ summary(.x)[["r.squared"]])) %>%
  arrange(rsq)
```

Let's use a function from broom to get the usual coefficient table from `summary.lm()` but in a friendlier form for downstream work.

```{r}
library(broom)
gap_fits %>%
  mutate(coef = map(fit, tidy)) %>%
  unnest(coef)
```
