---
title: "List columns"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Data frames are a fantastic data structure for data analysis. We usually think of them as a data receptacle for several atomic vectors with a common length and with a notion of "observation", i.e. the i-th value of each atomic vector is related to all the other i-th values.

But data frame are not limited to atomic vectors. They can host general vectors, i.e. *lists* as well. This is what I call a **list-column**.

List-columns and the data frame that hosts them require some special handling. In particular, it is highly advantageous if the data frame is a [tibble](https://github.com/tidyverse/tibble#readme), which anticipates list-columns. To work comfortably with list-columns, you need to develop techniques to:

  * **Inspect**. What have I created?
  * **Index**. How do I pull out specific bits by name or position?
  * **Compute**. How do I operate on my list-column to make another vector or list-column?
  * **Simplify**. How do I get rid of this list-column and back to a normal data frame?
  
The purrr package and all the techniques depicted in the other lessons come into heavy play here. This is a collection of worked examples that show these techniques applied specifically to list-columns.

## Regex and Trump tweets

### Load packages

```{r message = FALSE}
library(tidyverse)
library(stringr)
library(lubridate)
library(here) ## install.packages("krlmlr/here")
```

### Bring tweets in

Working with the same 7 tweets as [Trump Android words](ls08_trump-tweets.html) lesson. Go there for the rationale for choosing these 7 tweets.

```{r}
tb_raw <- read_csv(here("talks", "trump-tweets.csv"))
```

### Create a list-column of Trump Android words

Clean a variable and create a list-column:

  * `source` comes in an unfriendly form. Simplify to convey if tweet came from Android or iPhone.
  * `twords` are what we'll call the "Trump Android words". See [Trump Android words](ls08_trump-tweets.html) lesson for backstory. **This is a list-column!**

```{r}
source_regex <- "android|iphone"
tword_regex <- "badly|crazy|weak|spent|strong|dumb|joke|guns|funny|dead"

tb <- tb_raw %>%
  mutate(source = str_extract(source, source_regex),
         twords = str_extract_all(tweet, tword_regex))
```

### Derive new variables

Add variables, two of which are based on the `twords` list-column.

 * `n`: How many twords are in the tweet?
 * `hour`: At which hour of the day was the tweet?
 * `start`: Start character of each tword.
 
```{r}
tb <- tb %>%
  mutate(n = lengths(twords),
         hour = hour(created),
         start = gregexpr(tword_regex, tweet))
```

```{r include = FALSE}
# another possibilty that would require more processing
# so less useful for a talk example
# but more useful IRL:
# str_locate_all(tweet, tword_regex))
```

### Use regular data manipulation toolkit

Let's isolate tweets created before 2pm, containing 1 or 2 twords, in which there's an tword that starts within the first 30 characters.

```{r}
tb %>%
  filter(hour < 14,
         between(n, 1, 2),
         between(map_int(start, min), 0, 30))
```

Let's isolate tweets that contain both the twords "strong" and "weak".

```{r}
tb %>%
  filter(map_lgl(twords, ~ all(c("strong", "weak") %in% .x)))
```

## JSON from an API and Game of Thrones

## Nested data frame, modelling, and Gapminder


## Nesting data frames in a list column

*under development ... for now ...*

Head over to [Split-Apply-Combine in STAT 545](http://stat545.com/block024_group-nest-split-map.html)

o4

<http://r4ds.had.co.nz/many-models.html>

<!--

Remember Gapminder?

```{r}
library(gapminder)
gapminder %>% 
  ggplot(aes(year, lifeExp, group = country)) +
    geom_line(alpha = 1/3)
```

Get one meta-row per country

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()
by_country
by_country$data[[1]]
```

Compare contrast to a data frame grouped by country. *Or split on country?*.

Fit a model for each country.

```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}
models <- map(by_country$data, country_model)
by_country <- by_country %>% 
  mutate(model = map(data, country_model))
by_country
```

You can still use `filter()`, `arrange()`, etc.

```{r}
by_country %>% 
  filter(continent == "Europe")
by_country %>% 
  arrange(continent, country)
```

```{r}
library(broom)
by_country$model[[1]]
tidy(by_country$model[[1]])
glance(by_country$model[[1]])
augment(by_country$model[[1]])
```

-->
