---
title: "Nested data frames"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Load packages

Load purrr and repurrrsive, which contains recursive list examples. If you're just jumping here, the example datasets are introduced [elsewhere](https://jennybc.github.io/purrr-tutorial/ls00_inspect-explore.html), including via interactive listviewer widgets.

```{r}
library(tidyverse)
## install_github("jennybc/repurrrsive")
library(repurrrsive)
library(stringr)
```

## Lists as variables in a data frame

One row per GoT character. List columns for aliases and allegiances.

```{r}
x <- tibble(
  name = got_chars %>% map_chr("name"),
  aliases = got_chars %>% map("aliases"),
  allegiances = got_chars %>% map("allegiances")
)
#x %>% View()
```

What if we only care about characters with a "Lannister" alliance? Practice operating on a list-column.

```{r}
x %>%
  mutate(lannister = map(allegiances, str_detect, pattern = "Lannister"),
         lannister = map_lgl(lannister, any))
```

Keep only the Lannisters and Starks allegiance. You can use `filter()` with list-columns, but you will need to `map()` to list-ize your operation. Once I've got the characters I want, I drop `allegiances` and use `unnest()` to get back to a simple data frame with no list columns.

```{r}
x %>%
  filter(allegiances %>%
           map(str_detect, "Lannister|Stark") %>%
           map_lgl(any)) %>%
  select(-allegiances) %>%
  filter(lengths(aliases) > 0) %>%
  unnest() %>% 
  print(n = Inf)
```

```{r eval = FALSE, include = FALSE}
x_base <- data.frame(
  name = vapply(got_chars, `[[`, character(1), "name"),
  aliases = I(lapply(got_chars, `[[`, "aliases")),
  allegiances = I(lapply(got_chars, `[[`, "allegiances"))
)
keep1 <- vapply(x_base$allegiances, function(y) any(grepl("Lannister|Stark", y)), logical(1))
x_base <- x_base[keep1, ]
x_base$allegiances <- NULL
x_base
data.frame(
  name = rep(x_base$name, lengths(x_base$aliases)),
  aliases = unlist(x_base$aliases)
)
```

## Nesting data frames in a list column

*under development ... for now ...*

Head over to [Split-Apply-Combine in STAT 545](http://stat545.com/block024_group-nest-split-map.html)

o4

<http://r4ds.had.co.nz/many-models.html>

<!--

Remember Gapminder?

```{r}
library(gapminder)
gapminder %>% 
  ggplot(aes(year, lifeExp, group = country)) +
    geom_line(alpha = 1/3)
```

Get one meta-row per country

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()
by_country
by_country$data[[1]]
```

Compare contrast to a data frame grouped by country. *Or split on country?*.

Fit a model for each country.

```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}
models <- map(by_country$data, country_model)
by_country <- by_country %>% 
  mutate(model = map(data, country_model))
by_country
```

You can still use `filter()`, `arrange()`, etc.

```{r}
by_country %>% 
  filter(continent == "Europe")
by_country %>% 
  arrange(continent, country)
```

```{r}
library(broom)
by_country$model[[1]]
tidy(by_country$model[[1]])
glance(by_country$model[[1]])
augment(by_country$model[[1]])
```

-->
