---
title: "Gotchas"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

Mistakes I have made. And made. And made again. *Expect this page to grow as I (re)discover more gotchas.*

```{r}
library(purrr)
```

### The `magrittr` dot tension

The tidyverse takes its dot `.` pronoun from the [`magrittr` package](https://github.com/smbache/magrittr#readme). It means "the thing we are operating on" and is also known as the "argument placeholder".

You don't need the dot when you're using pipe-friendly functions and the planets align for you:

```{r}
8 %>% log2()
## is same as
log2(8)
```

But sometimes the thing you're passing into the right-hand side (RHS) is not the first argument:

```{r}
2 %>% log(8)
## is not what I want and is not the same as
2 %>% log(8, .)
## or 
2 %>% log(8, base = .)
```

And sometimes you want to prevent the left-hand side from being used as the (invisible) first argument on the RHS. So you have to enclose RHS in curly braces:

```{r}
iris %>% {
  c(rows = nrow(.), cols = ncol(.))
}
```

One last thing ... and this leads to the gotcha. The `.` can also be used to create a *unary function*:

```{r}
att <- . %>% toupper() %>% paste("ALL THE THINGS!")
"open source" %>% att()
"butter" %>% att()
"teach" %>% att()
```

What is `att` anyway?

```{r}
att
```

It is a "functional sequence".

It's fairly easy to write code where you think `.` is a placeholder, but it generates a functional sequence.

Watch me.

```{r}
library(purrr)
library(tibble)

x <- list(list(int = 1L, chr = 'a'), list(int = 2L, chr = 'b'))
  
## YES GOOD WORKS
x %>% {
  tibble(id = map_int(., "int"),
         chr = map_chr(., "chr"))
}

## NO BAD DOES NOT WORK
x %>% {
  tibble(id = . %>% map_int("int"),
         chr = . %>% map_chr("chr"))
}
```

What went wrong? 

`. %>% map_int("int")` built a unary function, instead of passing `x` into `map_int()`. Do not start a pipeline with `.` unless you want a unary function.

What does this have to do with `purrr`?

If you've got a complicated object `x` (e.g., a deeply nested list from JSON), you might build a data frame with repeated calls to `map_*()` functions. Be careful where you put your dot `.`!

### `purrr` is strict about types

`purrr`'s type checking is very strict, which is overhwhelmingly positive. But it will force you to be more aware of integer vs. double.

```{r error = TRUE}
set.seed(4561)
(x <- sample(1:5))

times_two <- function(x) x * 2
times_two(x)

x_list <- as.list(x)

## WTF?
x_list %>% 
  map_int(times_two)
```

Why can I suddenly not multiply these numbers by 2?

Because we've said to expect integers back and, though the elements of `x` are integer, the result of multiplying by the *double* 2 is *double*.

What can you do? Buckle down and make sure that integer stays integer, if that's appropriate. Or loosen up and use `map_dbl()` instead.

```{r}
## GOOD, in the buckle down sense
times_two <- function(x) x * 2L
x_list %>% 
  map_int(times_two)

## GOOD, in the loosen up sense
times_two <- function(x) x * 2
x_list %>% 
  map_dbl(times_two)
```

